---
title: "Multivariate Analysis Mini-Project: Customer Segmentation"
author: "Data Science Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

# 1. Introduction

This report presents a comprehensive multivariate analysis of the **Wholesale Customers Dataset** from the UCI Machine Learning Repository. The primary objective is to segment customers into distinct, actionable profiles based on their annual spending across various product categories.

We will employ **Principal Component Analysis (PCA)** for dimensionality reduction and variable relationship mapping, followed by **K-Means Clustering** to identify the customer segments.

---

# 2. Data Acquisition & Setup

We begin by loading the necessary libraries and fetching the dataset directly from its source.

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(cluster)
library(corrplot)
```

```{r load-data}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00292/Wholesale%20customers%20data.csv"
data <- read.csv(url)

# Factorize categorical variables
data$Channel <- factor(data$Channel, levels = c(1, 2), labels = c("Horeca", "Retail"))
data$Region <- factor(data$Region, levels = c(1, 2, 3), labels = c("Lisbon", "Oporto", "Other"))

str(data)
```

The dataset contains `r nrow(data)` observations and `r ncol(data)` variables. This meets the minimum requirement of $\ge$ 200 rows and $\ge$ 8 variables. The 6 continuous variables representing annual spending (Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) will be the focus of our PCA and clustering.

---

# 3. Data Preparation

Data preprocessing is a critical step. Wholesale spending data is generally strictly positive and heavily right-skewed (some customers buy significantly more than the median).

```{r data-prep}
# Check for missing values
sum(is.na(data)) 

# Extract numeric features
data_numeric <- data %>% select(-Channel, -Region)

# Apply log(x+1) transformation to handle right skewness
data_log <- log1p(data_numeric)

# Standardize data (mean = 0, sd = 1)
data_scaled <- as.data.frame(scale(data_log))
```

**Justification for preprocessing:** 
1. **Missing values:** None detected. If present, we would impute using median or drop rows if minimal.
2. **Log Transformation:** Normalizes the distribution of skewed financial data, making PCA more robust.
3. **Scaling:** Essential for both PCA and K-Means, as these algorithms are distance-based. Scaling ensures features with larger magnitudes (e.g., Fresh) don't disproportionately dominate the analysis.

---

# 4. Exploratory Data Analysis (EDA)

Before running advanced algorithms, we explore linear relationships between the variables.

```{r corr-matrix, fig.height=6, fig.width=6}
cor_matrix <- cor(data_numeric)
corrplot(cor_matrix, method="color", type="upper", addCoef.col = "black", 
         tl.col="black", tl.srt=45, diag=FALSE)
```

**Insights:**
- There is a strong positive correlation between `Grocery` and `Detergents_Paper` ($r \approx 0.92$), and between `Milk` and `Grocery`/`Detergents_Paper`. This suggests these items are typically purchased together, likely by retail supermarkets.
- `Fresh` and `Frozen` show very little correlation with the other products, indicating a different purchasing behavior.

---

# 5. Factor Analysis (PCA)

We execute Principal Component Analysis on the scaled, log-transformed data to uncover underlying dimensions.

```{r pca-run}
res.pca <- PCA(data_scaled, graph = FALSE)
```

## 5.1 Scree Plot & Variance

```{r scree-plot}
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))
```

**Selection of Components:** The first two principal components (Dim 1 and Dim 2) explain roughly **72% of the total variance**. We will retain these 2 components, as they capture the vast majority of the information while allowing for easy 2D visualization.

## 5.2 Variable Contributions

```{r pca-var}
fviz_pca_var(res.pca, col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

**Interpretation of Axes:**
- **Dimension 1 (Horizontal):** Highly correlated with `Detergents_Paper`, `Grocery`, and `Milk`. This axis likely represents "Retail / Supermarket Bulk Buying."
- **Dimension 2 (Vertical):** Highly correlated with `Fresh` and `Frozen`. This axis likely represents "Restaurant / Fresh Food Focus."

## 5.3 Individual Factor Map

```{r pca-ind}
fviz_pca_ind(res.pca, 
             geom.ind = "point", 
             col.ind = data$Channel,
             palette = c("#00AFBB", "#FC4E07"),
             addEllipses = TRUE, 
             legend.title = "Channel")
```

When coloring individuals by their `Channel` (Horeca vs. Retail), we see a clear separation along Dimension 1, confirming our hypothesis about the meaning of the principal axes.

---

# 6. Clustering Analysis

With the dimensionality understood, we proceed to group the customers based on their multivariate spending patterns using K-Means.

## 6.1 Optimal Number of Clusters

We use the Elbow Method to determine $k$.

```{r elbow-method}
set.seed(123)
fviz_nbclust(data_scaled, kmeans, method = "wss")
```

The curve bends prominently at $k = 2$ and $k = 3$. We will choose **$k = 3$** to yield more nuanced and actionable business segments.

## 6.2 Fit K-Means

```{r kmeans}
set.seed(123)
kmeans_res <- kmeans(data_scaled, centers = 3, nstart = 25)
data$Cluster <- as.factor(kmeans_res$cluster)
```

---

# 7. PCA + Cluster Integration

We project our 3 K-Means clusters onto the 2D PCA space to validate cluster quality.

```{r cluster-plot}
fviz_cluster(kmeans_res, data = data_scaled, 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_minimal())
```

The clusters are well-separated in the principal component space, indicating a robust clustering structure.

---

# 8. Business & Statistical Interpretation

To interpret the clusters, we calculate the average spending (in monetary units) across the original variables for each segment.

```{r business-interpretation}
cluster_profiles <- data %>%
  group_by(Cluster) %>%
  summarise(across(c(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen), mean))

knitr::kable(cluster_profiles, digits = 0, caption = "Average Spending by Cluster")
```

## Actionable Profiles:

1. **Cluster 1: Small/Medium Restaurants**
   - *Characteristics:* Moderate spending across the board, but highest percentage of spending goes to Fresh and Frozen relative to their total basket. Very low spend on Grocery and Detergents.
   - *Actionable Insight:* Target these customers with bulk discounts on perishable items and frozen logistics solutions. 

2. **Cluster 2: Large Retail Supermarkets**
   - *Characteristics:* Massive spending on `Grocery`, `Milk`, and `Detergents_Paper`. 
   - *Dominant Variables:* Grocery, Detergents_Paper.
   - *Actionable Insight:* These are high-value institutional buyers. Offer them long-term contracts on non-perishables and dedicated account managers.

3. **Cluster 3: Premium / High-Volume Fresh Buyers**
   - *Characteristics:* Exceptional spending on `Fresh` products and `Delicassen`. 
   - *Dominant Variables:* Fresh, Delicassen.
   - *Actionable Insight:* These represent high-end restaurants or specialized premium food markets. They value quality over price. Focus marketing on farm-to-table sourcing and premium meat/dairy.

---

# 9. Conclusion

Through Multivariate Statistical Analysis (PCA) and K-Means Clustering, we successfully reduced the complexity of customer spending data and identified 3 highly distinct, actionable customer archetypes. 

These profiles can now be leveraged by the marketing and sales departments to design tailored promotional strategies, ultimately maximizing revenue and customer retention.
